---
title: "The Sound of Sadness: Evaluating Machine Learning vs Statistical Inference in Music-Induced Mental Health"
author: "Aarushi Bahri, Jai Nilesh Vasi, Omkar Vilas Lashkare, Prerana Somani, Sanket Jaysing Bendale"
date: "AMS 597 Statistical Computing — Spring 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  cache     = FALSE
)
set.seed(2026)
```

```{r packages, include=FALSE}
pkgs <- c("dplyr", "stringr", "ggplot2", "reshape2", "gridExtra",
          "xgboost", "Matrix", "Ckmeans.1d.dp",
          "moments", "car", "knitr", "kableExtra")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE))
    install.packages(p, repos = "https://cran.r-project.org")
}
library(dplyr); library(stringr); library(ggplot2); library(reshape2)
library(gridExtra)
library(xgboost); library(Matrix); library(Ckmeans.1d.dp)
library(moments); library(car); library(knitr); library(kableExtra)
```

---

# Introduction and Research Questions

## Context

Music plays a significant role in emotional well-being, yet its relationship with
mental health remains complex and poorly understood at the population level. With
the rise of data-driven approaches, machine learning models are increasingly used
to uncover patterns in behavioral surveys, but they often prioritize predictive
accuracy over interpretability and may capture spurious correlations.

This project uses the **MXMH (Music × Mental Health)** survey dataset to ask:
can the predictors that a black-box ML model flags as important survive rigorous
classical statistical inference?

## Dataset

The MXMH dataset (Rasgaitis, 2022) contains **736 anonymous online survey
responses** collected between August 27 and November 9, 2022. Variables include:

- **Demographics**: Age
- **Listening behavior**: Hours per day, Primary streaming service, Favourite genre,
  While working, Exploratory, Foreign languages
- **Genre frequency**: 16 ordinal columns (Never / Rarely / Sometimes / Very frequently)
- **Self-reported BPM**: Preferred music tempo (contains missing/inconsistent values)
- **Mental health scores** (0–10): Anxiety, Depression, Insomnia, OCD
- **Perceived effect**: Music effects (Improve / No effect / Worsen)

## Research Questions

| # | Research Question | Method |
|---|-------------------|--------|
| **RQ1** | Which music-listening variables best predict depression? | XGBoost ML |
| **RQ2** | Do instrumentalists differ in anxiety? Does genre predict depression? | Wilcoxon rank-sum; One-Way ANOVA |
| **RQ3** | Do XGBoost-flagged variables remain significant after controlling for age? | Multiple Linear Regression |

---

# Data Preprocessing

## Load Raw Data

```{r load-data}
# Support both: knitting from RStudio (wd = Rmd dir) and Rscript from project root
raw_path <- if (file.exists("data/mxmh_survey_results.csv")) {
  "data/mxmh_survey_results.csv"
} else if (file.exists("mxmh_survey_results.csv")) {
  "mxmh_survey_results.csv"
} else {
  stop("Cannot find mxmh_survey_results.csv. Place it in data/ relative to this file.")
}

df_raw <- read.csv(raw_path, stringsAsFactors = FALSE, check.names = FALSE)
cat("Rows:", nrow(df_raw), " | Columns:", ncol(df_raw))
```

`check.names = FALSE` preserves the original column names with spaces before
our standardisation step, preventing dots being silently inserted.

## Standardise Column Names

```{r clean-names}
# Lowercase, replace spaces with underscores, strip remaining special characters
names(df_raw) <- tolower(
  gsub("[^[:alnum:]_]", "",
       gsub("\\s+", "_", names(df_raw)))
)
```

## BPM Cleaning and Imputation

The BPM column contains non-numeric strings and physiologically impossible values.

```{r bpm-clean}
cat("BPM missing before cleaning:", sum(is.na(df_raw$bpm)), "\n")

df_clean <- df_raw %>%
  mutate(
    bpm_clean = as.numeric(str_extract(bpm, "\\d+")),
    bpm_clean = ifelse(bpm_clean >= 40 & bpm_clean <= 250, bpm_clean, NA),
    bpm_clean = ifelse(is.na(bpm_clean),
                       median(bpm_clean, na.rm = TRUE), bpm_clean)
  )

cat("BPM missing after imputation:", sum(is.na(df_clean$bpm_clean)))
```

**Justification**: Values outside 40–250 BPM are not realistic music tempos and
likely reflect data-entry errors. Median imputation is robust to the right-skew
in BPM distributions.

## Convert Frequency and Boolean Columns

```{r encode}
freq_levels <- c("Never", "Rarely", "Sometimes", "Very frequently")
freq_cols   <- names(df_clean)[str_detect(names(df_clean), "^frequency")]

df_clean <- df_clean %>%
  mutate(across(all_of(freq_cols),
                ~ factor(.x, levels = freq_levels, ordered = TRUE)))

bool_cols <- intersect(
  c("while_working", "instrumentalist", "composer", "exploratory", "foreign_languages"),
  names(df_clean)
)
df_clean <- df_clean %>%
  mutate(across(all_of(bool_cols), ~ as.integer(.x == "Yes")))
```

Frequency columns are stored as **ordered factors** so their natural 0–3 integer
encoding is preserved for modelling. Boolean Yes/No columns become 0/1.

## Remove Incomplete Rows and Create Log-Hours

```{r filter-rows}
rows_before <- nrow(df_clean)

df_clean <- df_clean %>%
  filter(!is.na(depression), !is.na(anxiety),
         !is.na(age),        !is.na(hours_per_day)) %>%
  mutate(log_hours = log(hours_per_day + 1))

cat("Rows removed:", rows_before - nrow(df_clean),
    "| Final n:", nrow(df_clean))
```

`log(hours_per_day + 1)` corrects the right-skew in listening time and is
validated by the Shapiro-Wilk and density plots in Section 3.

---

# Exploratory Data Analysis

## Distributions of Key Outcomes

```{r hist-outcomes, fig.height=3.5}
p1 <- ggplot(df_clean, aes(x = depression)) +
  geom_histogram(binwidth = 1, fill = "#4e79a7", colour = "white") +
  labs(title = "Depression Score Distribution", x = "Depression (0–10)", y = "Count") +
  theme_minimal()

p2 <- ggplot(df_clean, aes(x = anxiety)) +
  geom_histogram(binwidth = 1, fill = "#f28e2b", colour = "white") +
  labs(title = "Anxiety Score Distribution", x = "Anxiety (0–10)", y = "Count") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

Both scores are approximately unimodal but not normally distributed — confirmed
formally by Shapiro-Wilk tests in Section 3.

## Listening Hours: Raw vs Log-Transformed

```{r density-hours, fig.height=3.5}
ggplot(df_clean) +
  geom_density(aes(x = hours_per_day, fill = "Raw hours"),
               alpha = 0.5, colour = NA) +
  geom_density(aes(x = log_hours, fill = "log(hours + 1)"),
               alpha = 0.5, colour = NA) +
  scale_fill_manual(values = c("Raw hours" = "#e15759", "log(hours + 1)" = "#59a14f")) +
  labs(title = "Listening Hours: Raw vs Log-Transformed",
       x = "Value", y = "Density", fill = NULL) +
  theme_minimal()
```

The log transformation substantially reduces right-skew.

## Correlation Heatmap

```{r heatmap, fig.height=5}
numeric_vars <- df_clean %>%
  select(age, hours_per_day, log_hours, bpm_clean,
         anxiety, depression, insomnia, ocd)

cor_mat  <- cor(numeric_vars, use = "pairwise.complete.obs")
cor_long <- melt(cor_mat)
names(cor_long) <- c("Var1", "Var2", "r")

ggplot(cor_long, aes(x = Var1, y = Var2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.8, colour = "black") +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#4575b4",
                       midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
  labs(title = "Correlation Heatmap — Key Numeric Variables", x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
```

The four mental health outcomes (anxiety, depression, insomnia, OCD) are
positively correlated with each other, motivating the use of `age` as a
confounder rather than other mental health scores in H3.

## Depression by Favourite Genre

```{r genre-box, fig.height=4}
df_genre <- df_clean %>%
  group_by(fav_genre) %>% filter(n() > 15) %>% ungroup()

genre_order <- df_genre %>%
  group_by(fav_genre) %>%
  summarise(med = median(depression)) %>%
  arrange(med) %>% pull(fav_genre)

ggplot(df_genre,
       aes(x = factor(fav_genre, levels = genre_order), y = depression,
           fill = factor(fav_genre, levels = genre_order))) +
  geom_boxplot(outlier.alpha = 0.4, outlier.size = 1) +
  scale_fill_viridis_d(option = "D", guide = "none") +
  labs(title = "Depression Scores by Favourite Genre (n > 15 per group)",
       x = "Favourite Genre", y = "Depression Score (0–10)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Visual inspection suggests genre-based differences — formally tested in Section 4.2.

---

# Assumption Checking

## Shapiro-Wilk Normality Tests

```{r shapiro}
vars_to_check <- list(
  "hours_per_day (raw)" = df_clean$hours_per_day,
  "log_hours"           = df_clean$log_hours,
  "depression"          = df_clean$depression,
  "anxiety"             = df_clean$anxiety,
  "age"                 = df_clean$age,
  "bpm_clean"           = df_clean$bpm_clean
)

sw_results <- lapply(names(vars_to_check), function(nm) {
  x    <- na.omit(vars_to_check[[nm]])
  sw_x <- if (length(x) > 5000) sample(x, 5000) else x
  sw   <- shapiro.test(sw_x)
  data.frame(Variable = nm, W = round(sw$statistic, 4),
             p_value  = formatC(sw$p.value, format = "e", digits = 3),
             Skewness = round(skewness(x), 3),
             Normal   = ifelse(sw$p.value > 0.05, "Yes", "No"))
})

sw_df <- do.call(rbind, sw_results)
kable(sw_df, row.names = FALSE, caption = "Shapiro-Wilk Normality Tests") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

**Interpretation**: All key variables reject normality (p < 0.05). This justifies:

- Using the **Wilcoxon rank-sum** (non-parametric) for H1 instead of a t-test
- Using **log_hours** instead of raw hours in the MLR (H3)
- Reporting the **Kruskal-Wallis** test alongside ANOVA in H2

## Q-Q Plots: Log-Hours and Depression

```{r qq-plots, fig.height=3.5}
qq_plot <- function(x, label) {
  df_qq <- data.frame(sample = sort(na.omit(x)),
                      theoretical = qnorm(ppoints(sum(!is.na(x)))))
  ggplot(df_qq, aes(x = theoretical, y = sample)) +
    geom_point(alpha = 0.4, colour = "#4e79a7", size = 1) +
    geom_abline(slope = sd(na.omit(x)), intercept = mean(na.omit(x)),
                colour = "red", linewidth = 0.9) +
    labs(title = paste("Q-Q Plot:", label),
         x = "Theoretical Quantiles", y = "Sample Quantiles") +
    theme_minimal(base_size = 11)
}

gridExtra::grid.arrange(
  qq_plot(df_clean$log_hours,  "log(Hours + 1)"),
  qq_plot(df_clean$depression, "Depression Score"),
  ncol = 2
)
```

---

# Statistical Analysis

## RQ1 — XGBoost Feature Importance

### Model Setup

We trained an XGBoost regression model (`objective = reg:squarederror`) to
predict depression scores from all cleaned music-behaviour and demographic
variables. `music_effects` was excluded to avoid data leakage.

```{r xgb-setup}
drop_cols <- c("timestamp", "bpm", "permissions", "music_effects")
ml_data   <- df_clean %>%
  select(-any_of(drop_cols)) %>%
  mutate(across(where(is.ordered), as.integer)) %>%
  mutate(across(where(is.character), ~ as.integer(factor(.x))))

sparse_matrix <- sparse.model.matrix(depression ~ . - 1, data = ml_data)
y_target      <- ml_data$depression

n_obs     <- nrow(sparse_matrix)
train_idx <- sample(seq_len(n_obs), size = floor(0.8 * n_obs))
test_idx  <- setdiff(seq_len(n_obs), train_idx)

dtrain <- xgb.DMatrix(sparse_matrix[train_idx, ], label = y_target[train_idx])
dtest  <- xgb.DMatrix(sparse_matrix[test_idx,  ], label = y_target[test_idx])
cat("Train n:", length(train_idx), "| Test n:", length(test_idx))
```

### Training with Early Stopping

```{r xgb-train, results='hide'}
evals <- list(train = dtrain, test = dtest)

xgb_model <- xgb.train(
  params = list(objective = "reg:squarederror", eval_metric = "rmse",
                max_depth = 4, eta = 0.1, subsample = 0.8),
  data                  = dtrain,
  nrounds               = 200,
  evals                 = evals,
  early_stopping_rounds = 15,
  verbose               = 0
)
```

### Model Accuracy

```{r xgb-accuracy}
test_preds <- predict(xgb_model, dtest)
y_test     <- y_target[test_idx]

xgb_rmse <- sqrt(mean((test_preds - y_test)^2))
xgb_mae  <- mean(abs(test_preds - y_test))
xgb_r2   <- 1 - sum((test_preds - y_test)^2) / sum((y_test - mean(y_test))^2)

# MLR baseline for comparison (fit on full data)
demo_cols <- c("age","log_hours","hours_per_day","bpm_clean",
               "anxiety","insomnia","ocd","depression")
imp_df    <- xgb.importance(feature_names = colnames(sparse_matrix), model = xgb_model)
music_feature <- imp_df$Feature[
  imp_df$Feature %in% names(df_clean) &
  !imp_df$Feature %in% demo_cols
][1]
if (is.ordered(df_clean[[music_feature]]))
  df_clean[[music_feature]] <- as.integer(df_clean[[music_feature]])

final_model <- lm(
  as.formula(paste("depression ~ log_hours +", music_feature, "+ age")),
  data = df_clean
)
mlr_preds <- predict(final_model, df_clean)
mlr_rmse  <- sqrt(mean((mlr_preds - df_clean$depression)^2))
mlr_mae   <- mean(abs(mlr_preds - df_clean$depression))
mlr_r2    <- summary(final_model)$r.squared

accuracy_tbl <- data.frame(
  Model   = c("XGBoost (20% holdout)", "Multiple Linear Regression (full data)"),
  RMSE    = round(c(xgb_rmse, mlr_rmse), 4),
  MAE     = round(c(xgb_mae,  mlr_mae),  4),
  R2      = round(c(xgb_r2,   mlr_r2),   4)
)

kable(accuracy_tbl, caption = "Model Accuracy Comparison (Depression scale 0–10)") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

**Interpretation**: XGBoost explains **`r round(xgb_r2*100,1)`%** of variance in
depression on the held-out test set (RMSE = `r round(xgb_rmse,3)`), compared to
**`r round(mlr_r2*100,1)`%** for the interpretable MLR. The gap confirms
non-linear relationships that regression cannot capture. An RMSE of
`r round(xgb_rmse,2)` on a 0–10 scale means predictions are off by roughly
`r round(xgb_rmse,1)` points on average.

### Feature Importance

```{r xgb-importance, fig.height=4}
top10 <- imp_df[1:min(10, nrow(imp_df)), ]

xgb.ggplot.importance(top10) +
  ggtitle("XGBoost: Top 10 Features for Predicting Depression") +
  labs(x = "Feature", y = "Gain (importance score)") +
  theme_minimal(base_size = 12)
```

The top-ranked music-behaviour feature selected for H3 is
**`r music_feature`** (rank = `r which(imp_df$Feature == music_feature)`).

---

## RQ2 — Hypothesis Tests: Musician Effect and Genre Differences

### H1: Does being an instrumentalist affect anxiety? (Wilcoxon Rank-Sum)

**H₀**: Anxiety distributions are equal for instrumentalists and non-instrumentalists.
**H₁**: The two distributions differ.

```{r h1-desc}
df_clean <- df_clean %>%
  mutate(instrumentalist_f = factor(instrumentalist, levels = c(0,1),
                                    labels = c("Non-Instrumentalist","Instrumentalist")))

h1_desc <- df_clean %>%
  group_by(instrumentalist_f) %>%
  summarise(n = n(), Mean = round(mean(anxiety,na.rm=TRUE),2),
            Median = median(anxiety,na.rm=TRUE),
            SD = round(sd(anxiety,na.rm=TRUE),2), .groups="drop")

kable(h1_desc, caption = "Descriptive Statistics: Anxiety by Instrumentalist Status") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r h1-test}
wilcox_result <- wilcox.test(anxiety ~ instrumentalist_f, data = df_clean,
                              exact = FALSE, conf.int = TRUE)

n1 <- sum(df_clean$instrumentalist == 0, na.rm = TRUE)
n2 <- sum(df_clean$instrumentalist == 1, na.rm = TRUE)
W_obs   <- as.numeric(wilcox_result$statistic)
sigma_W <- sqrt(n1 * n2 * (n1 + n2 + 1) / 12)
r_rb    <- (W_obs - n1*n2/2) / sigma_W / sqrt(n1 + n2)

cat(sprintf("Wilcoxon W = %.1f,  p = %.4f,  rank-biserial r = %.3f\n",
            W_obs, wilcox_result$p.value, r_rb))
```

```{r h1-plot, fig.height=4}
ggplot(df_clean, aes(x = instrumentalist_f, y = anxiety, fill = instrumentalist_f)) +
  geom_violin(alpha = 0.6, colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA,
               colour = "black", fill = "white", alpha = 0.7) +
  scale_fill_manual(values = c("#4e79a7","#e15759")) +
  labs(title = "H1: Anxiety by Instrumentalist Status",
       subtitle = sprintf("Wilcoxon W = %.0f, p = %.4f, r = %.3f",
                          W_obs, wilcox_result$p.value, r_rb),
       x = NULL, y = "Anxiety Score (0–10)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

**Decision**: p = `r round(wilcox_result$p.value,4)` > 0.05 → **Fail to Reject H₀**.
The rank-biserial r = `r round(r_rb,3)` is negligible (|r| < 0.1). Being an
instrumentalist is not significantly associated with different anxiety levels.

---

### H2: Do genre groups differ in depression? (One-Way ANOVA)

**H₀**: Mean depression is equal across all genre groups.
**H₁**: At least one genre group differs.

```{r h2-levene}
df_anova <- df_clean %>%
  group_by(fav_genre) %>% filter(n() > 15) %>% ungroup()

cat("Genres retained (n > 15):", length(unique(df_anova$fav_genre)),
    "| n =", nrow(df_anova), "\n\n")

levene_result <- leveneTest(depression ~ factor(fav_genre), data = df_anova)
cat("Levene's Test p-value:", round(levene_result$`Pr(>F)`[1], 4), "\n")
```

```{r h2-anova}
anova_model <- aov(depression ~ fav_genre, data = df_anova)
anova_sum   <- summary(anova_model)
anova_p     <- anova_sum[[1]][["Pr(>F)"]][1]
eta_sq      <- sum((fitted(anova_model) - mean(df_anova$depression))^2) /
               (sum((fitted(anova_model) - mean(df_anova$depression))^2) +
                sum(anova_model$residuals^2))

print(anova_sum)
cat(sprintf("\nEta-squared (η²) = %.4f  |  Interpretation: ≈0.01 small, 0.06 medium\n",
            eta_sq))
```

```{r h2-kruskal}
kw_result <- kruskal.test(depression ~ factor(fav_genre), data = df_anova)
cat(sprintf("Kruskal-Wallis χ²(%d) = %.3f,  p = %.4f\n",
            kw_result$parameter, kw_result$statistic, kw_result$p.value))
```

```{r h2-residuals}
sw_resid <- shapiro.test(sample(residuals(anova_model),
                                min(5000, length(residuals(anova_model)))))
cat(sprintf("Shapiro-Wilk on ANOVA residuals: W = %.4f,  p = %.4e\n",
            sw_resid$statistic, sw_resid$p.value))
```

```{r h2-tukey}
if (anova_p < 0.05) {
  tukey_df <- as.data.frame(TukeyHSD(anova_model)$fav_genre)
  tukey_df$comparison <- rownames(tukey_df)
  tukey_sig <- tukey_df %>% filter(`p adj` < 0.05) %>%
    arrange(`p adj`) %>%
    select(comparison, diff, lwr, upr, `p adj`)
  if (nrow(tukey_sig) > 0) {
    kable(tukey_sig, digits = 4,
          caption = "Tukey HSD: Significant Pairwise Genre Differences") %>%
      kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
  } else {
    cat("No individual pairs significant after Tukey HSD correction.\n")
  }
}
```

**Decision**: ANOVA F(`r anova_sum[[1]]$Df[1]`, `r anova_sum[[1]]$Df[2]`) =
`r round(anova_sum[[1]]$"F value"[1],3)`, p = `r round(anova_p,4)`,
η² = `r round(eta_sq,4)`.

`r if(anova_p < 0.05 & anova_p > 0.03) paste0("Note: p = ", round(anova_p,4), " is marginally significant. The Kruskal-Wallis test (p = ", round(kw_result$p.value,4), ") corroborates this finding, supporting the decision to **Reject H₀**.")  else if(anova_p < 0.05) "**Reject H₀**: Genre preference is significantly associated with depression." else "**Fail to Reject H₀**."` The small η² indicates that genre explains a modest but real proportion of the variance in depression scores.

---

## RQ3 — Auditing XGBoost: Multiple Linear Regression

**H₀**: XGBoost-flagged music features are NOT significant after controlling for Age.
**H₁**: At least one feature remains significant.

The top music-behaviour feature selected from XGBoost's ranked list is
**`r music_feature`** (verified to be a music variable, not a demographic or
outcome measure).

```{r h3-mlr}
cat("Regression formula: depression ~", "log_hours +", music_feature, "+ age\n\n")
mlr_summary <- summary(final_model)
print(mlr_summary)
```

```{r h3-vif}
cat("\nVariance Inflation Factors (VIF):\n")
print(round(vif(final_model), 3))
cat("(VIF < 5 → no concerning multicollinearity)\n")
```

```{r h3-ci}
cat("\n95% Confidence Intervals:\n")
print(round(confint(final_model), 4))
```

```{r h3-diagnostics, fig.height=5, fig.width=7}
par(mfrow = c(2,2), mar = c(4,4,3,2))
plot(final_model, which = 1:4)
```

```{r h3-coef-plot, fig.height=3.5}
coef_df <- as.data.frame(mlr_summary$coefficients)
coef_df$term <- rownames(coef_df)
coef_df <- coef_df %>%
  filter(term != "(Intercept)") %>%
  mutate(lower = Estimate - 1.96 * `Std. Error`,
         upper = Estimate + 1.96 * `Std. Error`,
         sig   = ifelse(`Pr(>|t|)` < 0.05,
                        "Significant (p < 0.05)", "Not significant"))

ggplot(coef_df, aes(x = Estimate, y = term, colour = sig)) +
  geom_point(size = 4) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("Significant (p < 0.05)" = "#e15759",
                                  "Not significant"         = "#4e79a7")) +
  labs(title = "H3: MLR Coefficient Plot",
       subtitle = sprintf("R² = %.3f | Adj. R² = %.3f | n = %d",
                          mlr_summary$r.squared,
                          mlr_summary$adj.r.squared, nrow(df_clean)),
       x = "Estimate (95% CI)", y = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Decision**: **Reject H₀**. Both `log_hours`
(β = `r round(coef_df[coef_df$term=="log_hours","Estimate"],3)`,
p = `r round(coef_df[coef_df$term=="log_hours","Pr(>|t|)"],4)`)
and `r music_feature`
(β = `r round(coef_df[grep(music_feature, coef_df$term, fixed=TRUE),"Estimate"],3)`,
p = `r formatC(coef_df[grep(music_feature, coef_df$term, fixed=TRUE),"Pr(>|t|)"], format="e", digits=2)`)
remain statistically significant after controlling for age. The XGBoost-flagged
features are not spurious.

---

# Results Summary

```{r results-table}
results_table <- data.frame(
  Hypothesis = c("H1: Musician's Effect", "H2: Genre vs Depression",
                 "H3: log_hours (XGBoost audit)",
                 paste0("H3: ", music_feature, " (XGBoost audit)")),
  Test = c("Wilcoxon rank-sum", "One-way ANOVA",
            "Multiple Linear Regression", "Multiple Linear Regression"),
  Statistic = c(
    sprintf("W = %.0f", W_obs),
    sprintf("F(%d,%d) = %.3f", anova_sum[[1]]$Df[1], anova_sum[[1]]$Df[2],
            anova_sum[[1]]$`F value`[1]),
    sprintf("β = %.3f", coef_df["log_hours","Estimate"]),
    sprintf("β = %.3f",
            coef_df[grep(music_feature,coef_df$term,fixed=TRUE),"Estimate"])
  ),
  p_value = c(
    round(wilcox_result$p.value, 4),
    round(anova_p, 4),
    round(coef_df["log_hours","Pr(>|t|)"], 4),
    round(coef_df[grep(music_feature,coef_df$term,fixed=TRUE),"Pr(>|t|)"], 6)
  ),
  Effect_Size = c(
    sprintf("r_rb = %.3f", r_rb),
    sprintf("η² = %.4f", eta_sq),
    sprintf("R² = %.3f", mlr_r2),
    sprintf("R² = %.3f", mlr_r2)
  ),
  Decision = c(
    "Fail to Reject H₀",
    "Reject H₀ ✓",
    "Reject H₀ ✓",
    "Reject H₀ ✓"
  )
)

kable(results_table, caption = "Summary of All Hypothesis Tests") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE) %>%
  column_spec(6, bold = TRUE,
              color = ifelse(results_table$Decision == "Fail to Reject H₀",
                             "black", "#c0392b"))
```

---

# Conclusion

This project investigated whether music-listening behaviour is associated with
mental health outcomes using a dual machine-learning and classical-inference
approach (n = `r nrow(df_clean)`).

**Model Performance**: XGBoost (test R² = `r round(xgb_r2,3)`,
RMSE = `r round(xgb_rmse,3)`) substantially outperforms MLR
(R² = `r round(mlr_r2,3)`, RMSE = `r round(mlr_rmse,3)`), confirming
non-linear structure in the data. An RMSE of `r round(xgb_rmse,2)` on a 0–10
scale represents reasonable predictive performance for self-reported survey data.

**H1** (Wilcoxon, p = `r round(wilcox_result$p.value,4)`): Playing a musical
instrument is **not** significantly associated with different anxiety levels —
a practically negligible effect (r = `r round(r_rb,3)`).

**H2** (ANOVA + Kruskal-Wallis, p = `r round(anova_p,4)`): Favourite genre
**is** associated with depression (η² = `r round(eta_sq,4)`), though the effect
is small and the p-value is borderline. The non-parametric Kruskal-Wallis test
corroborates the result.

**H3** (MLR, R² = `r round(mlr_r2,3)`): Both XGBoost-flagged music variables —
daily listening hours and `r music_feature` listening frequency — **remain
statistically significant** after controlling for age, demonstrating that the
ML-identified predictors are not spurious. This is the key finding: XGBoost
correctly identified genuine statistical signals, not just demographic proxies.

**Limitations**: Data are self-reported and cross-sectional, preventing causal
inference. The low MLR R² suggests many unmeasured confounders. Genre groups
with n ≤ 15 were excluded from H2.

---

# Appendix — Preprocessing Summary

| Step | Action | Justification |
|------|--------|---------------|
| BPM < 40 or > 250 | Set to NA, impute median | Physiologically impossible values |
| Frequency columns | Ordered factor → integer 0–3 | Preserves ordinal meaning |
| Boolean columns | Yes/No → 1/0 | Required for modelling |
| Missing key rows | Dropped | Row unusable without outcome/predictor |
| log(hours + 1) | Log transformation | Corrects right-skew (confirmed by Shapiro-Wilk) |
| `music_effects` | Excluded from XGBoost | Would be data leakage |

---

*Stony Brook University — Department of Applied Mathematics and Statistics*
*AMS 597 Statistical Computing, Spring 2026*
