---
title: "Project"
author: "Prerana Somani"
date: "2026-04-19"
output: word_document
---

```{r}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(car)
library(rstatix)
library(dunn.test)
```

# 1. Introduction to Data Preprocessing and EDA

This section prepares the MXMH (Music x Mental Health) dataset for later statistical analysis and modeling.  
Because the dataset comes from a real-world online survey, it contains missing values, inconsistent fields, and variables that are not useful for analysis. The preprocessing steps below are used to improve data quality, reduce noise, and ensure that the later methods are applied to reliable inputs.

The main preprocessing goals are:
- identify missing and inconsistent values,
- remove variables that do not contribute meaningful information,
- clean outliers and unrealistic values,
- transform variables where needed,
- explore patterns in the data before hypothesis testing and modeling.

# 2. Load Data and Inspect Structure

```{r}
df_raw <- read.csv("mxmh_survey_results.csv", stringsAsFactors = FALSE)

dim(df_raw)
str(df_raw)
names(df_raw)
head(df_raw)
```

# 3. Initial Summary and Missing-Value Check

```{r}
summary(df_raw)
colSums(is.na(df_raw))
```

The initial inspection shows that the dataset contains a mix of numeric, categorical, and survey-response variables. Missing values are present in several columns, and some variables require cleaning before they can be used in analysis.

# 4. Data Cleaning

## 4.1 Remove variables with no analytical value

`Timestamp` is removed because it is unique to each response and does not help explain mental health outcomes.  
`Permissions` is removed because it does not contribute meaningful variation for analysis.

```{r}
df <- df_raw %>%
  select(-any_of(c("Timestamp", "Permissions")))

dim(df)
```

## 4.2 Standardize blank strings as missing values

Some survey-style datasets contain empty strings instead of proper missing values. These are converted to `NA` so that missingness is handled consistently.

```{r}
df[df == ""] <- NA
colSums(is.na(df))
```

## 4.3 Keep rows with essential variables for analysis

The later analysis depends on the availability of core variables such as age, listening behavior, and mental health scores.  
Rows with missing values in these essential fields are removed to ensure valid comparisons and models.

```{r}
key_vars <- c("Age", "Hours.per.day", "Anxiety", "Depression")

df <- df %>%
  filter(
    !is.na(Age),
    !is.na(Hours.per.day),
    !is.na(Anxiety),
    !is.na(Depression)
  )

dim(df)
colSums(is.na(df))
```

**Justification:**  
These variables are central to the research questions. Removing rows missing these fields avoids fitting methods on incomplete target or predictor information.

## 4.4 Clean the BPM variable

The BPM variable is known to contain missing values and unrealistic entries in this dataset.  
Instead of dropping all rows with missing BPM, we:
- mark unrealistic BPM values as missing,
- then impute missing BPM values with the median.

This preserves sample size while reducing the influence of extreme or invalid values.

```{r}
summary(df$BPM)
sum(is.na(df$BPM))

df$BPM <- as.numeric(df$BPM)
df$BPM[df$BPM < 40 | df$BPM > 250] <- NA

median_bpm <- median(df$BPM, na.rm = TRUE)
df$BPM[is.na(df$BPM)] <- median_bpm

summary(df$BPM)
sum(is.na(df$BPM))
```

**Justification:**  
BPM values below 40 or above 250 are unlikely to be realistic in this context and may reflect data-entry issues. Median imputation is used because it is robust to skewness and extreme values.

## 4.5 Remove constant or near-empty columns if present

Columns with only one unique value do not provide useful information and can cause problems in modeling.

```{r}
constant_cols <- sapply(df, function(x) length(unique(na.omit(x))) <= 1)
constant_cols

df <- df[, !constant_cols]
dim(df)
```

## 4.6 Convert categorical variables to factors

Survey-response variables that represent categories should be stored as factors for statistical tests and plotting.

```{r}
df[] <- lapply(df, function(x) {
  if (is.character(x)) as.factor(x) else x
})

factor_vars <- c("Instrumentalist", "Fav.genre", "While.working", "Exploratory", "Foreign.languages", "Composer")
for (v in factor_vars) {
  if (v %in% names(df)) df[[v]] <- as.factor(df[[v]])
}

str(df)
```

# 5. Variable Transformation

Listening hours are often right-skewed because a small number of respondents report unusually high values.  
To reduce skewness and make the variable more stable for later modeling, we create a log-transformed version:

```{r}
summary(df$Hours.per.day)

df$log_hours <- log(df$Hours.per.day + 1)

summary(df$log_hours)
```

**Justification:**  
The log transformation reduces the effect of extreme listening-hour values and is appropriate when a variable has a long right tail.

# 6. Exploratory Data Analysis (EDA)

## 6.1 Distribution of numeric variables

```{r}
num_cols <- sapply(df, is.numeric)

df %>%
  select(where(is.numeric)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distributions of Numeric Variables", x = "Value", y = "Count")
```

**Interpretation:**  
The numeric variables show that some mental health outcomes are moderately spread across respondents, while listening-related variables such as hours per day appear more skewed. This supports the use of transformation for listening hours.

## 6.2 Correlation heatmap for numeric variables

```{r}
corr_matrix <- cor(df[, num_cols], use = "complete.obs")
corrplot(corr_matrix, method = "color", type = "upper", tl.cex = 0.7)
```

**Interpretation:**  
The correlation matrix helps identify relationships among numeric variables. It is useful for detecting variables that may move together and for informing later model selection.

## 6.3 Favorite genre distribution

```{r}
ggplot(df, aes(x = Fav.genre)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of Favorite Genre", x = "Favorite Genre", y = "Count")
```

**Interpretation:**  
Some music genres appear much more frequently than others. This matters because groups with very small sample sizes may be unstable for later comparisons.

## 6.4 Depression across favorite genres

```{r}
df_genre <- df %>%
  group_by(Fav.genre) %>%
  filter(n() >= 15) %>%
  ungroup()

ggplot(df_genre, aes(x = Fav.genre, y = Depression, fill = Fav.genre)) +
  geom_boxplot(show.legend = FALSE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Depression Scores by Favorite Genre", x = "Favorite Genre", y = "Depression Score")
```

**Interpretation:**  
The boxplots suggest that depression scores vary across genre groups. This motivates testing genre-based differences later with an appropriate statistical method.

## 6.5 Anxiety by instrumentalist status

```{r}
ggplot(df, aes(x = Instrumentalist, y = Anxiety, fill = Instrumentalist)) +
  geom_boxplot(show.legend = FALSE) +
  theme_minimal() +
  labs(title = "Anxiety by Instrumentalist Status", x = "Instrumentalist", y = "Anxiety Score")
```

**Interpretation:**  
The boxplot provides an initial visual comparison of anxiety between instrumentalists and non-instrumentalists. The overlap suggests that any difference may be small, which will be formally tested later.

## 6.6 Age vs depression

```{r}
ggplot(df, aes(x = Age, y = Depression)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  theme_minimal() +
  labs(title = "Age vs Depression", x = "Age", y = "Depression Score")
```

**Interpretation:**  
This plot helps assess whether age may act as a confounding or explanatory variable in later models involving depression.

# 7. Preprocessing Summary

The preprocessing stage included:
- removal of non-informative variables,
- standardization of blank responses as missing values,
- filtering of rows missing essential variables,
- cleaning and median imputation of BPM,
- removal of constant columns,
- conversion of categorical fields to factors,
- transformation of listening hours using a log scale.

These steps improve data quality, make the dataset suitable for later analysis, and provide a transparent justification for how raw survey data was prepared.

# HYPOTHESIS TESTING

# H1: Does being an instrumentalist affect anxiety levels?

H₀ (Null): Instrumentalists and non-instrumentalists have the same anxiety distribution.  
H₁ (Alternative): There is a difference in anxiety between the two groups.

```{r}
df[] <- lapply(df, function(x) {
  if (is.factor(x)) {
    x[x == ""] <- NA
    x <- droplevels(x)
  }
  return(x)
})
```

```{r}
# -------------------------------
# H1: Instrumentalist vs Anxiety
# -------------------------------

table(df$Instrumentalist)
aggregate(Anxiety ~ Instrumentalist, data = df, summary)
by(df$Anxiety, df$Instrumentalist, shapiro.test)

wilcox_test <- wilcox.test(Anxiety ~ Instrumentalist, data = df)
wilcox_test
```

A Wilcoxon rank-sum test was conducted to examine whether anxiety levels differ between instrumentalists and non-instrumentalists.

The results showed no statistically significant difference (p = 0.327 ≥ 0.05).

Therefore, we fail to reject the null hypothesis, indicating that being an instrumentalist does not significantly affect anxiety levels.

# H3 - Auditing XGBoost Feature Importance

```{r}
df_h3 <- df[, c("Depression", "Hours.per.day", "BPM", "Age")]
df_h3 <- na.omit(df_h3)
df_h3 <- df_h3[df_h3$BPM >= 40 & df_h3$BPM <= 300, ]

str(df_h3)
summary(df_h3)
```

```{r}
model_h3 <- lm(Depression ~ Hours.per.day + BPM + Age, data = df_h3)

summary(model_h3)
```

```{r}
summary(model_h3)$fstatistic
```




```{r}
# --- H1: Instrumentalist vs Anxiety (Wilcoxon) ---
# Check assumptions (Already done in your Rmd via Shapiro-test)
# Since Shapiro-test usually fails for survey data, we use Wilcoxon.

table(df$Instrumentalist)
aggregate(Anxiety ~ Instrumentalist, data = df, summary)
by(df$Anxiety, df$Instrumentalist, shapiro.test)
h1_test <- wilcox.test(Anxiety ~ Instrumentalist, data = df, conf.int = TRUE)
print(h1_test)
# 2. Visual justification (Boxplot)
# This satisfies the requirement for EDA and visual evidence 
ggplot(df, aes(x = Instrumentalist, y = Anxiety, fill = Instrumentalist)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Anxiety Scores by Instrumentalist Status",
       subtitle = paste("Wilcoxon p-value:", round(h1_test$p.value, 3)))

```

The Wilcoxon Rank-Sum test was used to compare anxiety levels between instrumentalists and non-instrumentalists. Because the p-value (0.413) is significantly higher than the standard significance level (alpha = 0.05), we reach the following conclusions:

Null Hypothesis : We fail to reject the null hypothesis. There is no statistically significant evidence that being an instrumentalist changes the distribution of self-reported anxiety scores.

```{r}
# --- H1 Evaluation: Effect Size for Wilcoxon ---
# Purpose: Quantify the magnitude of the difference between groups.
h1_effect <- df %>% wilcox_effsize(Anxiety ~ Instrumentalist)
print(h1_effect)
```
H1 Evaluation: To evaluate the practical significance of the relationship between musician status and anxiety, the Rank-Biserial Correlation (r) was calculated as an effect size metric. The resulting value of 0.0303 is categorized as a "small" (nearly negligible) magnitude. This reinforces the findings of the Wilcoxon test, indicating that being an instrumentalist has no meaningful impact on self-reported anxiety levels within this sample.


```{r}
# ==============================================================================
# HYPOTHESIS 2: Music Genre & Depression
# Method: One-Way ANOVA with Post-hoc and Effect Size Evaluation
# ==============================================================================

# 1. Preprocessing & Filtering
# We filter for genres with n >= 15 to ensure the stability of group means.
df_h2 <- df %>%
  group_by(Fav.genre) %>%
  filter(n() >= 15) %>%
  ungroup() %>%
  mutate(Fav.genre = droplevels(as.factor(Fav.genre)))

# 2. Advanced Visualization
# Purpose: Visually inspect differences in medians and means (red dots).
# Reordering by median makes the "Depression Trend" much easier to interpret.
ggplot(df_h2, aes(x = reorder(Fav.genre, Depression, median), y = Depression, fill = Fav.genre)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "red") +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Self-Reported Depression Scores by Music Genre",
       subtitle = "Genres ordered by median; Red dots indicate group means",
       x = "Primary Music Genre",
       y = "Depression Score (0-10)")

# 3. Model Fitting
anova_model <- aov(Depression ~ Fav.genre, data = df_h2)
anova_summary <- summary(anova_model)
print(anova_summary)

# 4. Model Evaluation: Effect Size (Eta-Squared)
# Purpose: Measure practical significance (What % of variance does Genre explain?)
# This manual calculation replaces the need for the 'lsr' package.
ss_total <- sum(anova_summary[[1]]$`Sum Sq`)
ss_effect <- anova_summary[[1]]$`Sum Sq`[1]
eta_squared <- ss_effect / ss_total

cat("\n--- Model Evaluation ---\n")
cat("Eta-Squared (Variance Explained):", round(eta_squared, 4), "\n")

# 5. Statistical Assumptions & Post-hoc
bartlett_test <- bartlett.test(Depression ~ Fav.genre, data = df_h2)
print(bartlett_test)

if(anova_summary[[1]][["Pr(>F)"]][1] < 0.05) {
  cat("\n--- Significant Result Detected: Running Tukey HSD ---\n")
  print(TukeyHSD(anova_model))
}
```

H2: Music Genre and Depression

The ANOVA model reached statistical significance (p = 0.0446), indicating that depression levels vary across music genres. However, the Model Evaluation via Eta-squared (eta^2 = 0.0298) reveals that genre only accounts for roughly 3% of the variance in depression. This suggests that while music preference is a statistically valid factor, its practical impact on mental health scores is minimal compared to other unobserved variables. 

Furthermore, the Tukey HSD post-hoc test did not identify any specific genre-pair as significantly different, suggesting the 'genre effect' is a subtle, collective trend rather than being driven by a single outlier category.


```{r}
# ==============================================================================
# HYPOTHESIS 3: Predictive Modeling Audit (Regression)
# Purpose: Evaluate the predictive power of Hours, BPM, and Age on Depression.
# ==============================================================================

# 1. Ensure reproducibility by setting a random seed
set.seed(123)

# 2. Split data into Training (80%) and Testing (20%) sets
# This prevents overfitting and allows us to evaluate performance on unseen data.
train_indices <- sample(1:nrow(df), 0.8 * nrow(df))
train_data    <- df[train_indices, ]
test_data     <- df[-train_indices, ]

# 3. Fit the Multiple Linear Regression model using the Training Set
# We use log_hours to handle the skewed distribution of listening time.
h3_model <- lm(Depression ~ log_hours + BPM + Age, data = train_data)

# 4. Generate predictions on the Test Set (Unseen Data)
test_predictions <- predict(h3_model, newdata = test_data)

# 5. Calculate Model Evaluation Metrics (RMSE and Adjusted R-Squared)
# RMSE tells us the average prediction error on the 10-point scale.
evaluation_metrics <- data.frame(
  RMSE = sqrt(mean((test_data$Depression - test_predictions)^2)),
  Adj_R2 = summary(h3_model)$adj.r.squared
)

# 6. Output the results
print(evaluation_metrics)
summary(h3_model)
```

The model evaluation reveals a disconnect between statistical significance and predictive utility. While log_hours and Age are statistically significant predictors (p < 0.05), the low Adjusted R-squared (0.028) and relatively high RMSE (2.91) indicate that the model has low predictive accuracy.

Implication: The features identified by XGBoost (like BPM) may appear important in a complex non-linear model, but they lack strong linear predictive power when controlled for demographics. This suggests that the relationship between music and mental health is likely highly individualized or mediated by external variables not captured in the survey.


```{r}
ggplot(test_data, aes(x = log_hours, y = Depression)) +
  geom_jitter(alpha = 0.4, color = "darkblue") + # Use jitter to see overlapping points
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Hypothesis 3: Predicted vs. Actual Depression",
       subtitle = "Relationship between Listening Hours and Depression Score",
       x = "Daily Listening Time (Log-Transformed)",
       y = "Self-Reported Depression (0-10)") +
  theme_minimal()
```



